Nilpotent Invariants \\ for Generic Discrete Series of Real Groups

### 2024-10-02 whittaker.tex
minor proofreading changes

We are interested in representations of~$G(\R)$ which admit Whittaker models. For this we consider pairs $(B,\eta)$ where 
\begin{enumerate}[(i)]
\item $B$ is a Borel subgroup of~$G$ defined over~$\R$, with real points $B(\R)$,
\item $\eta$ is a character of the nilpotent radical~$N(\R)$ of~$B(\R)$, subject to the condition that~$\eta$ is nontrivial on each simple root space.
\end{enumerate}
The existence of a Borel subgroup satisfying (i) is the definition of $G$ being quasisplit, which we assume holds from now on.

@@

There are other invariants attached to an
irreducible representation $\pi$.  We will be concerned with two of
them, defined very differently. The first is the \emph{wavefront set} $\WF(\pi)$,
which is an analytic invariant.
The second is the \emph{associated variety} $\AV(\pi)$, which is
defined  algebraically.
Both consist of nilpotent elements of the dual of the Lie algebra of~$G$, and are defined as
coadjoint orbits (of two different kinds).
The invariants $\Wh(\pi), \WF(\pi)$ and $\AV(\pi)$ are closely related to each other.

@@

The purpose
of this paper is to give a self-contained account of these matters. More precisely, we will do two things:
\begin{enumerate}
\item Give an elementary proof that  passage from~$\pi$ to $\mathfrak{w}(\pi)$, $\WF(\pi)$ and $\AV(\pi)$ defines bijections between generic discrete series in an $L$-packet for $G(\R)$, Whittaker data for $G(\R)$, and the appropriate sets of nilpotent orbits;
\item Given one of the invariants $\Wh(\pi)$, $\WF(\pi)$ or $\AV(\pi)$, explain in the simplest possible terms how to explicitly reconstruct the other two.
\end{enumerate}
Many results in this note are well known. We  tried to give new and simplified proofs for as many of them as we could; for instance, Section~\ref{sec:Sekiguchi} contains a simple proof that $\WF(\pi)$ and $\AV(\pi)$ are exchanged by the Kostant--Sekiguchi correspondence. Otherwise we tried to give convenient references.

@@

\subsubsection*{Nilpotent orbits in $\g^*$} 
\label{s:dual}
Let $\N$ be the set of nilpotent elements of $\g$. Several of the invariants below are related to~$\N$, but are more naturally defined as subsets of the dual vector space~$\g^\ast$.
Therefore we need to pass to the dual Lie algebra $\g^*$, and it is important to do this as canonically as possible, i.e. in a manner as independent as possible of a choice
of isomorphism $\g\simeq\g^*$.

@@

Using $\psi_\kappa$ we shall transfer to $\g^\ast$ several notions usually defined on $\g$: the asymptotic cone, the Kostant section, and the Kostant--Sekiguchi corresondence. In each case the resulting notions are independent of the
choice of $\kappa$. See Section~\ref{sec:Sekiguchi}.
In Section~\ref{s:AV} we need to be more precise: we choose  $\kappa$ to be the Killing form. (More generally, we could take any non-degenerate
invariant form which is negative definite on the Lie algebra of a maximal compact subgroup.)

@@

Suppose $\pi$ is an irreducible admissible representation of $G(\R)$  on a Hilbert space. 
The underlying
$(\g,K)$-module is on the space of $K$-finite vectors in
$V$; we will denote it by~$\pi_K$ in this paper.

@@

Now suppose $\pi$ is an irreducible admissible representation of $G(\R)$ on a Hilbert space $\mH$. 
The \emph{wave-front
  set} $\WF(\pi)$ is an analytic invariant determined by the
singularities of the global character of $\pi$ near the identity. See \cite{howe_wave_front, bv_local_structure, HarrisHeOlafsson} for the definition. 
The set~$\WF(\pi)$  is a closed $G(\R)$-invariant subset of the nilpotent cone in
$i\g(\R)^*$. As such it is the union of a finite number of nilpotent $G(\R)$-orbits. Again, this depends only on the equivalence class of~$\pi$.

@@

Suppose again that $\pi$ is an irreducible admissible representation of $G(\R)$ on a Hilbert space $\mH$. We consider the corresponding actions of $G(\R)$ on the space~$\mH^\infty$ of smooth vectors in $\mH$, and on its dual $\mH^{-\infty}$; and we still denote them by~$\pi$. Let~$(B, \eta)$ be a pair satisfying conditions (i)-(ii) of the Introduction. We say $\pi$ has a $(B,\eta)$-Whittaker model
if there exists a nonzero element $v$ in~$\mH^{-\infty}$ such that $\pi(n)(v)=\eta(n)v$ for all $n$ in the nilradical $N(\R)$ of $B(\R)$. See~\cite{matumoto}. If~$\w$ is the corresponding Whittaker datum, i.e. the $G(\R)$-conjugacy class of the pair $(B, \eta)$, we say~$\pi$ has a $\w$-Whittaker model.

@@

We now take~$A=G$, and continue to assume~$\tau$ is an involutive automorphism. Then the short exact sequence $1\rightarrow Z \rightarrow G \rightarrow \Gad\rightarrow 1$ 
gives rise to a ``long" exact sequence of pointed sets:
\begin{equation} \label{long_ptset}
1\rightarrow Z^\tau \rightarrow G^\tau \rightarrow \Gad^{\tau} \overset{\psi_\tau}{\longrightarrow} H^1(\tau,Z)\to H^1(\tau,G)\rightarrow H^1(\tau,\Gad).
\end{equation}
In~\eqref{long_ptset} the connecting map  $ \psi_\tau\colon \Gad^{\tau} \rightarrow H^1(\tau,Z)$ is defined as follows: if $\gamma \in \Gad^{\tau}$ is the coset $gZ$, then $g\tau(g^{-1}) \in Z^{-\tau}$, and $\psi_\tau$ sends~$\gamma$ to the class of $g \tau(g^{-1})$ in $H^1(\tau, Z)$. The kernel of~$\psi_\tau$ is $(G^\tau)_{\ad}$. Since~\eqref{long_ptset} is exact we deduce that $\psi_\tau$ induces a canonical  bijection
\begin{equation} \label{interp_q_cohomology} Q_{\tau}(G) \overset{\sim}{\longrightarrow}  \ker\left(H^1(\tau,Z)\overset{\varphi_\tau}{\longrightarrow} H^1(\tau,G)\right)\end{equation}
where $\varphi_\tau$ is induced by the inclusion $Z^{-\tau} \hookrightarrow G^{-\tau}$.

@@

@@

\subsection{Whittaker data of generic discrete series}

@@

Suppose $X\in i\g(\R)^*\subset \g^*$ is a regular nilpotent element (see Section \ref{s:dual}).
Let $\overline{\mathfrak b}(\R)$ be the kernel of $X$. This is the Lie algebra of~$\overline{B}(\R)$, where$\overline{B}$ is a Borel subgroup of~$G$ defined over $\R$.
Choose an $\R$-Cartan subgroup $T$ of
$\overline B$, and let~$B$ be the  Borel
subgroup opposite to~$\overline{B}$, characterized by $B\cap \overline B=T$.
Then $X$ defines a unitary
character of the nilradical  $N(\R)$  of $B(\R)$ by
\[ 
\psi_X(\exp Y)=e^{X(Y)}\quad (Y\in\n(\R)).
\] 
We define $\w_X$ to be the $G(\R)$-conjugacy class of $(B(\R),\psi_X)$; as the notation indicates this is independent 
of the choice of $T$.

@@

This confirms that  the corresponding statement  in the $p$-adic case \cite{debacker_reeder_generic, kaletha_epipelagic}
applies to the real case as well.

@@

We begin by defining different bases of~$\g$. Start with the following basis:
$$
E_\R=\begin{pmatrix}0&i\\0&0\end{pmatrix},\quad 
F_\R=\begin{pmatrix}0&0\\-i&0\end{pmatrix},\quad
H_\R=\begin{pmatrix}1&0\\0&-1\end{pmatrix}.
$$
These elements satisfy the standard $SL(2)$ commutation relations: 
\[ [E_\R, F_\R] = H_\R, \quad [H_\R, E_\R] = 2E_\R, \quad [H_\R, F_\R] = -2F_\R. \]

@@

First let us compute $\AC(G(\R)\cdot kH_\theta)=\AC(G(\R)\cdot\begin{pmatrix}0&-ik\\ik&0
\end{pmatrix})$ for $k\in \Z_{\ne 0}$.
Here we are taking the real asymptotic cone; since $kH_\theta\in i\g(\R)$ this is a subset of $i\g(\R)$. 
Taking the limit as $x\rightarrow \infty$ in 
$$
\begin{pmatrix}
x&0\\0&\frac 1x
\end{pmatrix}
\begin{pmatrix}0&-ik\\ik&0
\end{pmatrix}\begin{pmatrix}
\frac 1x&0\\0&x
\end{pmatrix}=
\begin{pmatrix}0&-ikx^2\\\frac{ik}{x^2}&0
\end{pmatrix}
$$
we see
$$
\AC(G(\R)\cdot kH_\theta)=-G(\R)\cdot\sgn(k)E_\R.
$$
Now we carry this computation over to $i\g(\R)^*$:
$$
\begin{aligned}
\AC(G(\R)\cdot\lambda_{ik})&=\AC(G(\R)\cdot kH_\theta^*)\\
&=\psi_c(\AC(G(\R)\cdot \psi_c\inv(kH_\theta^*))\\
&=\psi_c(-G(\R)\cdot\sgn(\frac k{2c})E_\R)\\
&=-G(\R)\cdot\sgn(\frac k{2c})cF_\R^*\\
&=-G(\R)\cdot\sgn(k)F_\R^*\\
&=G(\R)\cdot(-\sgn(k)E_\R^*).
\end{aligned}
$$
In the last line we used the fact that $E_\R^*,F_\R^*$ are $G(\R)$-conjugate.
Note that $c$ disappeared from the end result, as expected.

>>>>

We are interested in representations of~$G(\R)$ which admit Whittaker models. For this we consider pairs $(B,\eta)$ where 
\begin{enumerate}[(i)]
\item $B$ is a Borel subgroup of~$G$ defined over~$\R$,
\item $\eta$ is a character of the nilpotent radical~$N(\R)$ of~$B(\R)$, subject to the condition that~$\eta$ is nontrivial on each simple root space.
\end{enumerate}
The existence of a Borel subgroup satisfying (i) is the definition of $G$ being quasisplit, which we assume holds from now on.

@@

There are other invariants attached to an
irreducible representation $\pi$.  We will be concerned with two of
them, defined very differently. The first is the \emph{wavefront set} $\WF(\pi)$,
which is an analytic invariant.
The second is the \emph{associated variety} $\AV(\pi)$, which is
defined  algebraically.
Both consist of nilpotent elements of the dual of the Lie algebra of~$G$.
The invariants $\Wh(\pi), \WF(\pi)$ and $\AV(\pi)$ are closely related to each other.

@@

The purpose
of this paper is to give a self-contained account of these matters. More precisely, we will do two things:
\begin{enumerate}
\item Give an elementary proof that  passage from~$\pi$ to $\mathfrak{w}(\pi)$, $\WF(\pi)$ and $\AV(\pi)$ defines bijections between generic discrete series in an $L$-packet for $G(\R)$, Whittaker data for $G(\R)$, and the appropriate sets of nilpotent orbits;
\item Given one of the invariants $\Wh(\pi)$, $\WF(\pi)$ or $\AV(\pi)$, explain in the simplest possible terms how to explicitly reconstruct the other two.
\end{enumerate}
Many results in this note are well known. We give new and simplified proofs for as many of them as we could; for instance, Section~\ref{sec:Sekiguchi} contains a simple proof that $\WF(\pi)$ and $\AV(\pi)$ are exchanged by the Kostant--Sekiguchi correspondence. Otherwise we give convenient references.

@@

\subsubsection*{Nilpotent orbits in $\g^*$} 
\label{s:dual}
Let $\N$ be the set of nilpotent elements of $\g$. Several of the invariants below are related to~$\N$, but are more naturally defined as subsets of the dual vector space~$\g^\ast$.
Therefore we need to pass to the dual Lie algebra $\g^*$, and it is important to do this as independently as possible of a choice
of isomorphism $\g\simeq\g^*$.

@@

Using $\psi_\kappa$ we shall transfer to $\g^\ast$ several notions usually defined on $\g$: the asymptotic cone, the Kostant section, and the Kostant--Sekiguchi corresondence. In each case the resulting notions are independent of the
choice of $\kappa$. See Sections~\ref{s:gds} and \ref{sec:Sekiguchi}.
In Section~\ref{s:AV} we need to be more precise: we choose  $\kappa$ to be the Killing form  (or any non-degenerate
invariant form which is negative definite on the Lie algebra of a maximal compact subgroup.)

@@

Suppose $\pi$ is an irreducible admissible representation of $G(\R)$  on a Hilbert space. 
The underlying
$(\g,K)$-module is on the space of $K$-finite vectors in
$V$; we will denote it by~$\pi_K$.

@@

Now suppose $\pi$ is an irreducible admissible representation of $G(\R)$ on a Hilbert space $\mH$. 
The \emph{wave-front
  set} $\WF(\pi)$ is an analytic invariant determined by the
singularity of the global character of $\pi$ near the identity. See \cite{howe_wave_front, bv_local_structure, HarrisHeOlafsson} for the definition. 
The set~$\WF(\pi)$  is a closed $G(\R)$-invariant subset of the nilpotent cone in
$i\g(\R)^*$. As such it is the union of a finite number of nilpotent $G(\R)$-orbits. Again, this depends only on the equivalence class of~$\pi$.

@@

Suppose again that $\pi$ is an irreducible admissible representation of $G(\R)$ on a Hilbert space $\mH$. We consider the corresponding actions of $G(\R)$ on the space~$\mH^\infty$ of smooth vectors in $\mH$, and on its dual $\mH^{-\infty}$. Let~$(B, \eta)$ be a pair satisfying conditions (i)-(ii) of the Introduction. We say $\pi$ has a $(B,\eta)$-Whittaker model
if there exists a nonzero element $v$ in~$\mH^{-\infty}$ such that $\pi(n)(v)=\eta(n)v$ for all $n$ in the nilradical $N(\R)$ of $B(\R)$. See~\cite{matumoto}. If~$\w$ is the corresponding Whittaker datum, i.e. the $G(\R)$-conjugacy class of the pair $(B, \eta)$, we say~$\pi$ has a $\w$-Whittaker model.

@@

We now take~$A=G$, and continue to assume~$\tau$ is an involutive automorphism. Then the short exact sequence $1\rightarrow Z \rightarrow G \rightarrow \Gad\rightarrow 1$ 
gives rise to a long exact sequence of pointed sets:
\begin{equation} \label{long_ptset}
1\rightarrow Z^\tau \rightarrow G^\tau \rightarrow \Gad^{\tau} \overset{\psi_\tau}{\longrightarrow} H^1(\tau,Z)\to H^1(\tau,G)\rightarrow H^1(\tau,\Gad).
\end{equation}
In~\eqref{long_ptset} the connecting map  $ \psi_\tau\colon \Gad^{\tau} \rightarrow H^1(\tau,Z)$ is defined as follows: if $\gamma \in \Gad^{\tau}$ is the coset $gZ$, then $g\tau(g^{-1}) \in Z^{-\tau}$, and $\psi_\tau$ sends~$\gamma$ to the class of $g \tau(g^{-1})$ in $H^1(\tau, Z)$. The kernel of~$\psi_\tau$ is $(G^\tau)_{\ad}$. Since~\eqref{long_ptset} is exact we deduce that $\psi_\tau$ induces a canonical  bijection
\begin{equation} \label{interp_q_cohomology} Q_{\tau}(G) \overset{\sim}{\longrightarrow}  \ker\left(H^1(\tau,Z)\overset{\varphi_\tau}{\longrightarrow} H^1(\tau,G)\right)\end{equation}
where $\varphi_\tau$ is induced by the inclusion $Z^{-\tau} \hookrightarrow G^{-\tau}$.

@@

\qed

@@

\subsection{Whittaker data of generic discrete series}
\label{s:gds}

@@

Suppose $X\in i\g(\R)^*\subset \g^*$ is a regular nilpotent element (see Section \ref{s:dual}).
Let $\overline{\mathfrak b}(\R)$ be the kernel of $X$. This is the Lie algebra of~$\overline{B}(\R)$, where
$\overline{B}$ is an $\R$-Borel subgroup of~$G$.
Choose an $\R$-Cartan subgroup $T$ of
$\overline B$, and let~$B$ be the  Borel
subgroup opposite to~$\overline{B}$, characterized by $B\cap \overline B=T$.
Then $X$ defines a unitary
character of the nilradical  $N(\R)$  of $B(\R)$ by
\[ 
\psi_X(\exp Y)=e^{X(Y)}\quad (Y\in\n(\R)).
\] 
We define $\w_X$ to be the $G(\R)$-conjugacy class of $(B(\R),\psi_X)$; as the notation indicates this is independent 
of the choice of $T$.

@@

@@

We begin by defining two  bases of~$\g$. Start with the following basis:
$$
E_\R=\begin{pmatrix}0&i\\0&0\end{pmatrix},\quad 
F_\R=\begin{pmatrix}0&0\\-i&0\end{pmatrix},\quad
H_\R=\begin{pmatrix}1&0\\0&-1\end{pmatrix}.
$$
These elements satisfy the standard $SL(2)$ commutation relations: 
\[ [E_\R, F_\R] = H_\R, \quad [H_\R, E_\R] = 2E_\R, \quad [H_\R, F_\R] = -2F_\R. \]

@@

First we compute $\AC(G(\R)\cdot kH_\theta)=\AC(G(\R)\cdot\begin{pmatrix}0&-ik\\ik&0
\end{pmatrix})$ for $k\in \Z_{\ne 0}$.
Here we are taking the real asymptotic cone; since $kH_\theta\in i\g(\R)$ this is a subset of $i\g(\R)$. 
Taking the limit as $x\rightarrow \infty$ in 
$$
\begin{pmatrix}
x&0\\0&\frac 1x
\end{pmatrix}
\begin{pmatrix}0&-ik\\ik&0
\end{pmatrix}\begin{pmatrix}
\frac 1x&0\\0&x
\end{pmatrix}=
\begin{pmatrix}0&-ikx^2\\\frac{ik}{x^2}&0
\end{pmatrix}
$$
we see
$$
\AC(G(\R)\cdot kH_\theta)=-G(\R)\cdot\sgn(k)E_\R.
$$
Now we carry this computation over to $i\g(\R)^*$:
$$
\begin{aligned}
\AC(G(\R)\cdot\lambda_{ik})&=\AC(G(\R)\cdot kH_\theta^*)\\
&=\psi_c(\AC(G(\R)\cdot \psi_c\inv(kH_\theta^*))\\
&=\psi_c(-G(\R)\cdot\sgn(\frac k{2c})E_\R)\\
&=-G(\R)\cdot\sgn(\frac k{2c})cF_\R^*\\
&=-G(\R)\cdot\sgn(k)F_\R^*\\
&=G(\R)\cdot(-\sgn(k)E_\R^*).
\end{aligned}
$$
In the last line we used the fact that $E_\R^*,F_\R^*$ are $G(\R)$-conjugate.
Note that $c$ disappeared from the end result, as expected.
